Formation of porous crystals via viscoelastic phase separation

### 2016-12-19 nature/crystal_gel_r.tex
Add surface particles: - draw them in dark purple - differentiate from liquid in evolution of fraction of particles - sketch a surface layer around the hexagon sketching the crystal

\begin{figure}
 \includegraphics[width=10cm]{network}
 % crystal_size.eps: 0x0 pixel, 300dpi, 0.00x0.00 cm, bb=(atend)
\caption{{\bf Percolated network structures.} {\bf a:} gel; {\bf b:} crystal-gel. The gel state is obtained from the state
point $\phi\approx 0.33$ and $c_p=1.07$ mg/g at $t=400$ min after sample preparation, while the crystal-gel is obtained at the same $\phi$ but
at lower polymer concentration, $c_p=0.38$ mg/g at $t=468$ min after sample preparation. On the left, the network is represented in a coarse-grained fashion (see Methods). 
We can clearly see not only the percolated nature of both networks, but also the difference in the smoothness of the network surface between amorphous and crystal gels. 
On the right, slabs of 4.5 $\mu$m thickness and 41 $\mu$m width are plotted, with particles coloured with the same colour scheme as in Fig.~\ref{fig:transitions}: red (crystal), green (gas particles), white (liquid particles). 
We note that very few liquid particles remain in \textbf{b} and the network is almost crystalline. 
} 
\label{fig:network}
\end{figure}

A direct analysis of the colloidal positions at late times confirms that indeed the new structure obtained at
$\phi\approx 0.33$ and $c_p=0.38$ mg/g has different morphological properties than usual colloidal gels.
We directly compare the two structures in Fig.~\ref{fig:network}, where Gaussian filtering is used to depict the network structure as a continuous field (see Methods).
Panel a presents the results for high
polymer concentration gel, $c_p=1.07$ mg/g; panel b presents the results for the low polymer concentration sample, $c_p=0.38$ mg/g.
The comparison shows that the new arrest mechanism produces a network structure with bigger pores, smoother surfaces \textcolor{blue}{(consistently with $D=3$)}, and thicker strands (panel b) compared
to the colloidal gel sample (panel a). The right column shows the individual particle positions for slabs with a thickness of 5 particle diameters. 
From the figure it is immediately clear that the network strands at low polymer concentration are crystalline. The new arrest mechanism thus
involve the formation of a crystal-gel network. In order to explain the morphological differences between colloidal gels and crystal-gels,
we now consider the kinetics of formation in our samples.

@@

\begin{figure*}
 \centering
 \includegraphics[width=12cm]{transitions}
 % snapshots.eps: 0x0 pixel, 300dpi, 0.00x0.00 cm, bb=0 -1 2256 735
 \caption{{\bf Crystal gel formation.}  
{\bf a,} Reconstructions from confocal coordinates at $\phi\approx 0.33$ and $c_p=0.38$ mg/g. The depth of view is 9.0 $\mu$m, while the lateral dimension is 41 $\mu$m. Particles are drawn to scale and coloured according to their phase. Red: crystal; white: liquid; green: gas.
{\bf b,} Fraction of particles for each phase. Gray vertical lines indicate the times shown in panel a. 
{\bf c,} Each particle during its trajectory changes between three states: gas, liquid and crystal. Direct crystallization (continuous arrow) accounts for 54\% of particles trajectories in which
gas or liquid particles transition to the crystal state without liquid evaporation or crystal de-sublimation. The Bergeron process (dotted arrow) accounts for 35\% of particle trajectories in which liquid particles
transition to the gas state before crystallizing. Ostwald ripening (dashed arrow) accounts for 11\% of particle trajectories in which crystal particle transition to the gas state before returning
to the crystal state.}
\label{fig:transitions}
\end{figure*}

>>>>

\begin{figure}
 \includegraphics[width=10cm]{network}
 % crystal_size.eps: 0x0 pixel, 300dpi, 0.00x0.00 cm, bb=(atend)
\caption{{\bf Percolated network structures.} 
{\bf a, b:} the network is represented in a coarse-grained fashion (see Methods).
{\bf c, d:} slabs of 4.5 $\mu$m thickness and 41 $\mu$m width. Particles are drawn to scale and coloured according to their phase. Purple: crystal; dark purple: crystal surface; gray: liquid; orange: gas. 
{\bf a, c:} a gel obtained from the state point $\phi\approx 0.33$ and $c_p=1.07$ mg/g at $t=400$ min after sample preparation. 
{\bf b, d:} a crystal-gel at the same $\phi$ but at lower polymer concentration, $c_p=0.38$ mg/g at $t=468$ min after sample preparation. 
On the top row we can clearly see not only the percolated nature of both networks, but also the difference in the smoothness of the network surface between amorphous and crystal gels. 
We note that very few liquid particles remain in \textbf{d} and the network is almost crystalline. 
} 
\label{fig:network}
\end{figure}

A direct analysis of the colloidal positions at late times confirms that indeed the new structure obtained at $\phi\approx 0.33$ and $c_p=0.38$ mg/g has different morphological properties than usual colloidal gels.
We directly compare the two structures in Fig.~\ref{fig:network}a and b, where Gaussian filtering is used to depict the network structure as a continuous field (see Methods).
Panel~a presents the results for high polymer concentration gel, $c_p=1.07$ mg/g \textcolor{blue}{that never undergoes stress-driven ageing}; panel~b presents the results for the low polymer concentration sample, $c_p=0.38$ mg/g.
The comparison shows that the new arrest mechanism produces a \textcolor{blue}{beaded} network structure with bigger pores, smoother surfaces \textcolor{blue}{(consistently with $D=3$)}, and thicker strands (panel b) compared 
to the colloidal gel sample (panel a). The bottom row shows the individual particle positions for slabs with a thickness of 5 particle diameters. 
From the figure it is immediately clear that the network strands at low polymer concentration are crystalline. The new arrest mechanism thus
involve the formation of a crystal-gel network. In order to explain the morphological differences between colloidal gels \textcolor{blue}{(networks made of thin strands)} and crystal-gels \textcolor{blue}{(networks made of thick crystal beads)},
we now consider the kinetics of formation in our samples.

@@

\begin{figure*}
 \centering
 \includegraphics[width=16cm]{transitions}
 % snapshots.eps: 0x0 pixel, 300dpi, 0.00x0.00 cm, bb=0 -1 2256 735
 \caption{{\bf Crystal gel formation.}  
{\bf a,} Reconstructions from confocal coordinates at $\phi\approx 0.33$ and $c_p=0.38$ mg/g. The depth of view is 9.0 $\mu$m, while the lateral dimension is 41 $\mu$m. Particles are drawn to scale and coloured according to their phase. Purple: crystal; dark purple: crystal surface; gray: liquid; orange: gas.
{\bf b,} Fraction of particles for each phase. Gray vertical lines indicate the times shown in panel a. 
{\bf c,} Sketch of the three routes by which a particle can end up in a crystal state. 
Probabilities are computed from the history of each single trajectory. 
A trajectory without liquid evaporation or crystal sublimation is counted as direct crystallization (continuous arrow, 54\%). 
If a liquid particle evaporates and then de-sublimates it counts in the Bergeron process (dotted arrow, 35\%). 
If sublimation is followed by de-sublimation it is Ostwald ripening (dashed arrow, 11\%). 
Most trajectories proceed via the surface state.}
\label{fig:transitions}
\end{figure*}

### 2016-12-19 nature/response3.tex
Add surface particles: - draw them in dark purple - differentiate from liquid in evolution of fraction of particles - sketch a surface layer around the hexagon sketching the crystal

\noindent
{\bf Dear Pep,}
\vskip 0.3cm
We would like to thank you for your providing us with two Referee reports for our manuscript (NM15113574). We understand your decision to reject the manuscript, but, after careful consideration of the Referee reports, we couldn't avoid thinking that they are mostly supportive of our work, and the only Referee (Referee \#1) that is somewhat against publication is basing his argument on a probable misunderstanding of the true novelty of the first part of our work.

We decided to write to you again to clarify that the Referee might have missed some very important points. In general, both Referees seem to have appreciated the improvements
we made after the first round of reviews: Reviewer \#1 calls the work \emph{very beautiful and interesting}, while Referee \#3 goes into the details of explaining how we actually satisfactorily clarified his previous doubts, with just a couple of minor points left. Reviewer \#2 was very positive during the first round, but unfortunately didn't submit a second report.
Overall we were able to find only two criticism in the reports.
\begin{itemize}
 \item Referee \#1 laments that the first part of the paper describes spinodal decomposition without showing anything new from previous studies (but he admits the novelty of the study of the different crystallisation channels). But this is not the case, as the crystallisation of the liquid network requires an additional step, called stress-driven aging, that was never observed experimentally before.
In our new version of the manuscript we spent considerable time developing a new analysis, that shows how spinodal decomposition proceeds differently depending on the polymer concentration, and that the aging of the network due to mechanical stress is a necessary condition for the subsequent crystallization process. We think that this completely invalidates the point of the Referee, because in addition to different crystallization channels, our study unveils a new link between the aging of the liquid network and the subsequent crystallization process.
 \item Referee \#3 suggests that we better acknowledge that the crystal-gel is found only in a quite specific region of the phase diagram, and this also became the main motivation for the rejection of our manuscript. We also fundamentally disagree with this idea. It's true that crystal-gel states don't appear everywhere in the phase diagram, but this doesn't mean that they are hard to obtain. Rather the opposite, there is a reproducible region around the critical point were these states can be consistently obtained, and we were able to reproduce these findings both in our new cell design, and in ordinary capillaries, showing that the region of crystal-gel formation is well within the experimental range. Moreover the extent of this region can be widened by changing the location of the critical point, which in our system is controlled by the length of the non-adsorbing polymer: moving the critical point to lower polymer concentrations opens the window where bonds can rearrange before the intervening glass transition. So we argue that the region of formation of crystal-gel porous structures is not only easily accessible, but also controllable.
\end{itemize}

We hope that you'll see that both criticisms are not indicative of any weaknesses in our work. We also took advantage of the new suggestions to improve the manuscript even further. In particular, concerning the point raised by Referee \#1 we have introduced a new type of analysis, which looks at the bond-breaking probability as a function of time, during the whole process of spinodal decomposition. Our new results show that, at low polymer-concentrations, the network undergoes a structural reorganisation, which results in more compact structures, that are a necessary condition for the appearance of the first crystalline nuclei. To our knowledge this is the first experimental demonstration of this process: it is not captured by computer simulations, as they generally neglect hydrodynamic interactions which play a fundamental role in this kinetic process, and no previous experiment has had the capability of tracking particles during the early stages of spinodal decomposition.

We are rather positive that the new version of the manuscript would put the Referees in a position to better understand the full extent of our results, and
so we hope that you might give it another chance. If that is is not the case, we are still grateful for the opportunity that the comments gave us for improving the manuscript in a meaningful way.

@@

We thank the Referee for her/his appreciation of the improvements made during our last revision of the manuscript. The Referee comments on the first 7 pages being not original, and simply describing the process of spinodal decomposition as previously observed in experiments and simulations. But we believe that our study offers novel insights also in the description of the early stages of the phase separation process. This is a very important point, as there is a clear link between the early-stage kinetics of the liquid network, and the late-stage crystallization process. In order to highlight this connection, in this new version of the manuscript we introduce a new type of analysis. In summary, we show that our work represents the first experimental demonstration of a process called \emph{stress-driven aging}. Most past simulation works have neglected, or not fully taken into account, the role of hydrodynamic interactions (HI) in the phase separation process. As was shows for example in Ref.~\cite{furukawa2010key}, the phase separation process without HI produces rather compact structures, which would then easily crystallize to form compact nuclei. This is not what happens if one take HI into consideration. Due to the incompressibility condition, the liquid network has a lower-connectivity, and the viscoelastic phase separation results in the accumulation of mechanical stress on these liquid bonds. To things can happen depending on the polymer concentration. At high-polymer concentration, the bonds are not-easily broken, and the network which is first formed cannot undergo structural reorganisation. This inevitably leads to gelation. On the other side, if the polymer concentration is low enough, the liquid network can reorganise with a rather sharp transition to a more compact network. This condition is necessary for the appearance of the first crystalline nuclei. In our new manuscript, we are able to follow the kinetics of these processes by looking at the time evolution of the bond-breaking probability. The different network states are in fact characterised by different bond-breaking probabilities and by different connectives. In Fig.~2a,b we show these quantities for high-polymer concentration samples, where spinodal decomposition is the only process that occurs. At low-polymer concentrations (Fig.~2c,d) instead, the network undergoes a sharp transition to a state with lower bond-breaking probability and higher connectivity. This transition is the stress-driven aging process, and increases the connectivity of the network. Despite being theoretically predicted, Ref.~\cite{tanaka2000,tanaka2007spontaneous}, this transition was not directly observed before.
We agree that the previous version of the manuscript was lacking on this part, so the Referee comments inspired us to develop the new analysis presented in Fig.2.

@@

We fully agree that the evidence from Fig.~1 c,d was not conclusive, but the goal of those figures was not to provide conclusive evidence, but to provide a comprehensive study of the early-stages of spinodal decomposition. In this context the Fig.~1 c,d only provide hints that the cp = 0.38 mg/g state point behaves differently from other state points. The rest of the manuscript and the Supplementary materials is devoted to providing the conclusive evidence that that is the case. We have made this more clear in the new manuscript.

\textcolor{red}{The Referee really likes Supplementary Fig3, so at this point if we want to answer him fully, we should probably replace Fig. 3c with Supplementary Figure 3. What do you think?}

@@

We also agree that our previous manuscript didn't say much about the actual kinetic mechanism by which crystal-gels form. This point was also raised by Referee \#1 in noting that we didn't provide any additional insight in the phase separation process. In order to fully address this issue, we have implemented a kinetic analysis of the trajectories at all state points, computing the probability of bond-breaking between consecutive frames. This new kinetic analysis reveals that at low-polymer concentrations a new mechanism (stress-driven aging) is taking place, which has the effect of increasing the average coordination of the colloidal particles in the liquid phase.
We believe this to be a novel observation of this phenomena. Most past simulation works have neglected, or not fully taken into account, the role of hydrodynamic interactions (HI) in the phase separation process. As was shows for example in Ref.~\cite{furukawa2010key}, the phase separation process without HI produces rather compact structures, which would then easily crystallize to form compact nuclei. This is not what happens if one take HI into consideration. Due to the incompressibility condition, the liquid network has a lower-connectivity, and the viscoelastic phase separation results in the accumulation of mechanical stress on these liquid bonds. To things can happen depending on the polymer concentration. At high-polymer concentration, the bonds are not-easily broken, and the network which is first formed cannot undergo structural reorganisation. This inevitably leads to gelation. On the other side, if the polymer concentration is low enough, the liquid network can reorganise with a rather sharp transition to a more compact network. This condition is necessary for the appearance of the first crystalline nuclei. In our new manuscript, we are able to follow the kinetics of these processes by looking at the time evolution of the bond-breaking probability. The different network states are in fact characterised by different bond-breaking probabilities and by different connectives. In Fig.~2a,b we show these quantities for high-polymer concentration samples, where spinodal decomposition is the only process that occurs. At low-polymer concentrations (Fig.~2c,d) instead, the network undergoes a sharp transition to a state with lower bond-breaking probability and higher connectivity. This transition is the stress-driven aging process, and increases the connectivity of the network. Despite being theoretically predicted, Ref.~\cite{tanaka2000,tanaka2007spontaneous}, this transition was not directly observed before.

>>>>

\noindent
{\bf Dear Pep and/or a newly assigned Editor in Nature Materials,}
\vskip 0.7cm
We would like to thank you for your providing us with two Referee reports for our manuscript (NM15113574). 
We are glad to see that both Referees 1 and 3 seem to have appreciated the improvements 
we made after the first round of reviews: Reviewer \#1 calls the work \emph{very beautiful and interesting}, while Referee \#3 goes into the details of explaining how we actually satisfactorily clarified her/his previous doubts, with just a couple of minor points left. Reviewer \#2 was very positive during the first round.

After receiving the comments of the Referees' reports, we have taken them seriously and made an effort to address them properly. 
We summarize how we have addressed the critical comments of the Referees.
\begin{itemize}
 \item Referee \#1 seemed to think that the first part of the paper describes spinodal decomposition without showing anything new from previous studies (but she/he admits the novelty of the study of the different crystallzation channels). 
We are afraid that there may be some misunderstanding here. 
This is because the crystallisation of the liquid network requires an additional step, called stress-driven ageing, 
that was never observed experimentally before. However, this criticism makes us realize that this point was not so clear 
in the previous manuscript. 
To improve this, we spent considerable time developing a new analysis, which shows how spinodal decomposition proceeds differently depending on the polymer concentration, and that the ageing of the network due to mechanical stress is a necessary condition for the subsequent crystallization process. We think that this has considerably increased the impact of our work, because in addition to different crystallization channels, this additional work unveils a new link between the ageing of the liquid network and the subsequent crystallization process.
\item Referee \#3 suggests that we better acknowledge that the crystal-gel is found only in a quite specific region of the phase diagram. 
It's certainly true that crystal-gel states don't appear everywhere in the phase diagram, but this doesn't mean that they are hard to obtain. 
The Referee's comment led us to consider this problem in more depth. 
First we already showed that there is a reproducible region around the critical point were these states can be consistently obtained, and we were able to reproduce these findings both in our new cell design, and in ordinary capillaries, showing that the region of crystal-gel formation is well within the experimental range. Moreover, thanks to the above-mentioned new analysis, we have successfully unveiled the physical requirements for crystal-gel observation. This mechanism suggests that the extent of this region can be widened by changing the location of the critical point, which in our system is controlled by the size of the non-adsorbing polymer: moving the critical point to lower polymer concentrations opens the window where bonds can rearrange before the intervening glass transition. So we may say that the region of formation of crystal-gel porous structures is not only easily accessible, but also controllable.
\end{itemize}

We hope that the above explanation have addressed the comments of the Referees satisfactorily. We also took advantage of the new suggestions to improve the manuscript even further. In particular, as mentioned above, we have introduced a new type of analysis, which looks at the bond-breaking probability as a function of time, during the whole process of spinodal decomposition. Our new results show that, at low polymer-concentrations, the network undergoes a structural reorganisation, which results in more compact structures, that are a necessary condition for the appearance of the first crystalline nuclei. To our knowledge this is the first experimental demonstration of this process: it is not captured by computer simulations, as they generally neglect hydrodynamic interactions which play a fundamental role in this kinetic process, and no previous experiment has had the capability of tracking particles during the early stages of spinodal decomposition.

\vskip 0.3cm

Since the manuscript now includes the new analysis mentioned above and is expanded with 6 figures, we would like to resubmit it as an Article rather than a Letter. 
\textit{
\vskip 0.3cm}

We hope that you and your reviewers will find that the revised manuscript is now suitable for publication in Nature Materials.

Warm thanks to you for your kind consideration and cooperation.

@@

We thank the Referee for her/his appreciation of the improvements made during our last revision of the manuscript. The Referee comments on the first 7 pages being not original, and simply describing the process of spinodal decomposition as previously observed in experiments and simulations. But we believe that our study offers novel insights also in the description of the early stages of the phase separation process. This is a very important point, as there is a clear link between the early-stage kinetics of the liquid network, and the late-stage crystallization process. In order to highlight this connection, in this new version of the manuscript we introduce a new type of analysis. In summary, we show that our work represents the first experimental demonstration of a process called \emph{stress-driven aging}. Most past simulation works have neglected, or not fully taken into account, the role of hydrodynamic interactions (HI) in the phase separation process. As was shown for example in Ref.~\cite{
furukawa2010key}, the phase separation process without HI produces rather compact structures, which would then easily crystallize to form compact nuclei. This is not what happens if one takes HI into consideration. Due to the incompressibility condition, the liquid network has a lower-connectivity, and the viscoelastic phase separation results in the accumulation of mechanical stress on these liquid bonds. Two things can happen depending on the polymer concentration. At high-polymer concentration, the bonds are not-easily broken, and the network which is first formed cannot undergo structural reorganisation. This inevitably leads to gelation. On the other side, if the polymer concentration is low enough, the liquid network can reorganise with a rather sharp transition to a more compact network. This condition is necessary for the appearance of the first crystalline nuclei. In our new manuscript, we are able to follow the kinetics of these processes by looking at the time evolution of the bond-breaking 
probability. The different network states are in fact characterised by different bond-breaking probabilities and by different coordination numbers. In Fig.~2a,b we show these quantities for high-polymer concentration samples, where spinodal decomposition is the only process that occurs. At low-polymer concentrations (Fig.~2c,d) instead, the network undergoes a sharp transition to a state with lower bond-breaking probability and higher connectivity. This transition is the stress-driven aging process, and increases the connectivity of the network. Despite being theoretically predicted, Ref.~\cite{tanaka2000,tanaka2007spontaneous}, this transition was not directly observed before.
We agree that the previous version of the manuscript was lacking on this part, so the Referee comments inspired us to develop the new analysis presented in new Fig. 2.

@@

We fully agree that the evidence from Fig.~1 c,d was not conclusive, but the goal of those figures was not to provide conclusive evidence, but to provide a comprehensive study of the early-stages of spinodal decomposition. In this context the Fig.~1 c,d only provide hints that the cp = 0.38 mg/g state point behaves differently from other state points. Direct evidence comes from the use of bond-orientational order parameters, presented in Fig. 5 and 6. 
Following the Referee's kind suggestion, we have now moved previous Fig. S3 to a new Fig. 3b. 
We hope that this will make the difference of the cp = 0.38 mg/mg state point from the others clearer.

@@

We also agree that our previous manuscript didn't say much about the actual kinetic mechanism by which crystal-gels form. This point was also raised by Referee \#1 in noting that we didn't provide any additional insight in the phase separation process. In order to fully address this issue, we have implemented a kinetic analysis of the trajectories at all state points, computing the probability of bond-breaking between consecutive frames. This new kinetic analysis reveals that at low-polymer concentrations a new mechanism (stress-driven aging) is taking place, which has the effect of increasing the average coordination of the colloidal particles in the liquid phase.
We believe this to be a novel observation of this phenomenon. Most past simulation works have neglected, or not fully taken into account, the role of hydrodynamic interactions (HI) in the phase separation process. As was shows for example in Ref.~\cite{furukawa2010key}, the phase separation process without HI produces rather compact structures, which would then easily crystallize to form compact nuclei. This is not what happens if one take HI into consideration. Due to the incompressibility condition, the liquid network has a lower-connectivity, and the viscoelastic phase separation results in the accumulation of mechanical stress on these liquid bonds. Two things can happen depending on the polymer concentration. At high-polymer concentration, the bonds are not-easily broken, and the network which is first formed cannot undergo structural reorganisation. This inevitably leads to gelation. On the other side, if the polymer concentration is low enough, the liquid network can reorganise with a rather sharp 
transition to a more compact network. This condition is necessary for the appearance of the first crystalline nuclei. In our new manuscript, we are able to follow the kinetics of these processes by looking at the time evolution of the bond-breaking probability. The different network states are in fact characterised by different bond-breaking probabilities and by different coordination numbers. In Fig.~2a,b we show these quantities for high-polymer concentration samples, where spinodal decomposition is the only process that occurs. At low-polymer concentrations (Fig.~2c,d) instead, the network undergoes a sharp transition to a state with lower bond-breaking probability and higher connectivity. This transition is the stress-driven aging process, and increases the connectivity of the network. Despite being theoretically predicted, Ref.~\cite{tanaka2000,tanaka2007spontaneous}, this transition was not directly observed before in experiments. 
We stress that for the experimental observation we need to follow particles at a single-particle level from the beginning of phase separation. This was made possible by a new experimental protocol of salt injection combined with confocal microscopy observation.
